---
title: "Far-right voting in Flanders 1987-2024"
author: "Jonas Vanhoef"
date: "2026-06-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# 1. Introduction

This document describes the preparation and integration of election results and geographical data used in the subsequent statistical analyses.
The full breakdown of the data used, methodology and computational procedure can be found in the thesis.



# 2. Required packages

The following packages are used for data import, manipulation, spatial processing, visualisation, and statistical modelling.

```{r, warning=FALSE, message=FALSE}
library(xml2)
library(dplyr)
library(purrr)
library(stringr)
library(sf)
library(ggplot2)
library(tidyr)
library(tmap)
library(readxl)
library(lme4)
library(purrr)
library(performance)
library(broom.mixed)
```



# 3. Imports

The datasets used through this analysis.

```{r warning=FALSE, message=FALSE}
#geographical data
cantons_geom <- st_read("geographical_data/territorialdivisions.gpkg")
muni_geom <- st_read("geographical_data/stat_sectors_2024.sqlite")

#election data
files_can <- list.files("election_data", pattern = "\\.xml$", full.names = TRUE)
files_muni <- list("election_data/2014.xml", "election_data/2019.xml", "election_data/2024.xml")

#canton and municipal merger look-up tables
muni_merged_lookup <- read_excel("geographical_data/municipal_mergers.xlsx")
canton_lookup <- read_excel("geographical_data/Kantons_gemeenten.xlsx")

#independent variables
file_inhabitants <- "variables/inhabitants.xlsx"
file_imm_noneu <- "variables/non_eu_immigrants.xlsx"
imi_mergers <- read.csv("variables/non_eu_immigration.csv")
prosperity <- read_excel("variables/prosperity_index.xlsx", sheet = "pros_index_fl")
crime <- read_excel("variables/crime.xlsx")

#election years
election_years <- c(1987, 1991, 1995, 1999, 2003, 2007, 2010, 2014, 2019, 2024)
```



# 4. Canton-level elction results
# 4.1. Extraction of election results

Extract canton-level results from the XML election files (canton name, number of valid votes and number of votes for VB).

```{r warning=FALSE, message=FALSE}
#read every file and create one dataframe per year
vb_results_can <- map(files_can, function(file){

  #year from filename
  year <- str_extract(basename(file), "\\d{4}")
  
  #read XML
  doc <- read_xml(file)

  #select canton nodes
  cantons <- xml_find_all(doc, ".//level[@description='Canton']")

  #extract canton results
  results <- map_dfr(cantons, function(canton){

    canton_name <- xml_text(xml_find_first(canton, "./labels/label[@language='nl']"))
    valid_votes <- as.numeric(xml_attr(canton, "valid-votes"))

    #find Vlaams Belang or Vlaams Blok party
    vb <- xml_find_first(canton, "./lists/list[@party='VLAAMS BELANG' or @party='VLAAMS BLOK']")

    if (inherits(vb, "xml_missing")) {
      vb_votes <- 0
    } else {
      vb_votes <- as.numeric(xml_attr(vb, "votes"))
    }

    tibble(
      year = as.integer(year),
      canton = canton_name,
      vb_votes = vb_votes,
      valid_votes = valid_votes
    )
  })

  results
})


#clean results
vb_results_can <- bind_rows(vb_results_can)
vb_results_can$canton <- sub("^Kanton\\s+", "", vb_results_can$canton)
```



# 4.2. Identification of Flemish cantons

The election files contain cantons from across Belgium. The following step therefore excludes cantons that do not belong to the Flemish analytical sample.
In addition, historical canton names are harmonised to their corresponding later names where geographical changes or municipal mergers require this.

```{r warning=FALSE, message=FALSE}
#create list of non-Flemish cantons and filter them
cantons_non_fl <- list("Aarlen", "Aat", "Andenne", "Anderlecht", "Anderlues", "Antoing", "Aubel", "Aywaille", "Bastenaken", 
                       "Beaumont", "Beauraing", "Beloeil", "Bergen", "Binche", "Bitsingen", "Borgworm", "Bouillon", "Boussu",
                       "Brussel", "Brussel (E5)", "BZ Waals-Brabant", "BzAntwerpen", "BZBrussel-Halle-Vilvoorde", "BZHenegouwen",
                       "BZLeuven", "BZLimburg", "BZLuik", "BZLuxemburg", "BZNamen", "BZOost-Vlaandren", "BZWest-Vlaanderen", 
                       "Celles", "Charleroi", "Châtelet", "Chièvres", "Chimay", "Ciney", "Couvin", "Dinant", "Dison", "Doornik", 
                       "Dour", "Doornik", "Durbuy", "E5-Kieskring Antwerpen", "E5-Kieskring Henegouwen", "E5-Kieskring Limburg",
                       "E5-Kieskring Luik", "E5-Kieskring Luxemburg", "E5-Kieskring Namen", "E5-Kieskring Oost-Vlaanderen", 
                       "E5-Kieskring Vlaams-Brabant", "E5-Kieskring Waals-Brabant", "E5-Kieskring West-Vlaanderen", "Edingen", 
                       "Eghezée", "Elsene", "Erezée", "Estaimpuis", "Etalle", "Eupen", "Fauvillers", "Ferrières", "Fléron", 
                       "Florennes", "Florenville", "Fontaine-l'Evêque", "Fosses-la-Ville", "Frameries", "Frasnes-lez-Anvaing", 
                       "Gedinne", "Geldenaken", "Gembloux", "Genepiën", "Grâce-Hollogne", "Hannuit", "Héron", "Herstal", "Herve", 
                       "Hoei", "Houffalize", "Komen-Waasten", "La Louvière", "La Roche-en-Ardenne", "Le Roeulx", "Lens", "Lessen", 
                       "Leuze-en-Hainaut", "Limburg", "Luik", "Malmedy", "Marche-en-Famenne", "Merbes-le-Château", "Messancy", 
                       "Moeskroen", "Namen", "Nandrin", "Nassogne", "Neufchâteau", "Nijvel", "Paliseul", "Péruwelz", "Perwijs", 
                       "Philippeville", "Pont-à-Celles", "Rochefort", "Saint-Hubert", "Saint-Nicolas", "Saint-Vith", "Sainte-Ode", 
                       "Sankt-Vith", "Schaarbeek", "Seneffe", "Seraing", "Sint-Gillis", "Sint-Jans-Molenbeek", 
                       "Sint-Joost-ten-Node", "Spa", "Speciaal Kanton: Buza Antwerpen", "Speciaal Kanton: Buza Brussel-Hoofdstad", 
                       "Speciaal Kanton: Buza Brussel-Hoofdstad (E3,E4,E5)", "Speciaal Kanton: Buza Henegouwen", 
                       "Speciaal Kanton: Buza Leuven", "Speciaal Kanton: Buza Limburg", "Speciaal Kanton: Buza Luik", 
                       "Speciaal Kanton: Buza Luxemburg", "Speciaal Kanton: Buza Namen", "Speciaal Kanton: Buza Oost-Vlaanderen", 
                       "Speciaal Kanton: Buza Waals Brabant", "Speciaal Kanton: Buza West-Vlaanderen", "Speciaal: BuZa Antwerpen", 
                       "Speciaal: BuZa Brussel-Halle-Vilvoorde", "Speciaal: BuZa Henegouwen", "Speciaal: BuZa Leuven", 
                       "Speciaal: BuZa Limburg", "Speciaal: BuZa Luik", "Speciaal: BuZa Luxemburg", "Speciaal: BuZa Namen", 
                       "Speciaal: BuZa Oost-Vlaanderen", "Speciaal: Buza Waals Brabant", "Speciaal: BuZa West-Vlaanderen", 
                       "Stavelot", "Stemmen Uitgebracht In het Kanton Sint-Genesius-Rode", "Thuin", "Ukkel", "Verlaine", "Verviers",
                       "Vielsalm", "Virton", "Vloesberg", "Walcourt", "Waver", "Wellin", "Wezet", "Zinnik", 
                       "Speciaal: Buza Waals Brabant ", "Stemmen Uitgebracht In Het Kanton Sint-Genesius-Rode", 
                       "Sint-Joost-Ten-Node", "Frasnes-Lez-Anvaing", "Étalle")


vb_results_can <- vb_results_can %>% filter(!canton %in% cantons_non_fl)

#account for name changes in cantons over time
vb_results_can$canton <- sub("Nevele", "Aalter", vb_results_can$canton)
vb_results_can$canton <- sub("Oosterzele", "Merelbeke", vb_results_can$canton)
vb_results_can$canton <- sub("Kruishoutem", "Kruisem", vb_results_can$canton)
vb_results_can$canton <- sub("Harelbeke", "Waregem", vb_results_can$canton)
vb_results_can$canton <- sub("^Puurs$", "Puurs-Sint-Amands", vb_results_can$canton)
vb_results_can$canton <- sub("Neerpelt", "Pelt", vb_results_can$canton)
vb_results_can$canton <- sub("Mesen", "Heuvelland", vb_results_can$canton)
vb_results_can$canton <- sub("Beveren-Waas", "Beveren", vb_results_can$canton)

#remove Sint-Genesius-Rode and Zaventem due to incomplete number of elections
vb_results_can <- vb_results_can %>% filter(!canton %in% "Sint-Genesius-Rode")
vb_results_can <- vb_results_can %>% filter(!canton %in% "Zaventem")

#merge cantons due to changes over time
vb_results_can <- vb_results_can %>%
  mutate(canton = case_when(canton == "Zomergem" | canton == "Waarschoot" ~ "Lievegem",
                            canton == "Lichtervelde" ~ "Hooglede",
                            canton == "Zonnebeke" ~ "Ieper",
                            TRUE ~ canton)) %>%
  group_by(year, canton) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop")

#calculate vote share
vb_results_can <- vb_results_can %>% mutate(vb_perc = (vb_votes/valid_votes)*100) 
vb_results_can$vb_perc <- round(vb_results_can$vb_perc, 2)
```



# 4.3. Vlaams Belang vote share over time

The following figure provides a descriptive overview of the VB vote share at Flemish scale over the observation period.

```{r warning=FALSE, message=FALSE}
#calculate percentage of votes for the Chamber as compared to votes for Flanders
tot_perc_vb <- vb_results_can %>% group_by(year) %>% summarise(tot_perc = (sum(vb_votes)/sum(valid_votes)*100))

#plot Flanders-wide results
ggplot(tot_perc_vb, aes(year, tot_perc, label = round(tot_perc, 2))) + geom_line() + geom_point() +
  scale_x_continuous(
    breaks = c(1987, 1991, 1995, 1999, 2003, 2007, 2010, 2014, 2019, 2024),
    limits = c(1987, 2026)) +
  scale_y_continuous(breaks = c(0, 5, 10, 15, 20, 25), limits = c(1, 24)) +
  theme_classic() + xlab("Election year") + ylab("% votes VB") + geom_text(nudge_y = -0.3, nudge_x = 1.5, size = 3.5)

#clean up
vb_results_can <- vb_results_can %>% select(-vb_votes, -valid_votes)
```


# 4.4. Preparation of spatial election data

The cleaned canton-level election results are joined to the geographical boundaries. The resulting dataset is reshaped into a wide format, with one column per election year, and exported as a GeoPackage for subsequent spatial analysis and visualisation.

```{r warning=FALSE, message=FALSE}
#merge election data and geographical data
cantons_geom <- cantons_geom %>% select(cantonheadoffice, shape)
cantons_full <- vb_results_can %>% left_join(cantons_geom, by = c("canton" = "cantonheadoffice"))

#pivot table
cantons_plot <- cantons_full %>% select(canton, year, vb_perc, shape)
cantons_plot <- cantons_plot %>%
  pivot_wider(names_from = year, values_from = vb_perc)

#export data
st_write(cantons_plot, "output/cantons_plot.gpkg")
```



# 5. Municipal-level elction results
# 5.1. Extraction of election results

Extract Municipal-level results from the XML election files (Municipality name, number of valid votes and number of votes for VB).

```{r warning=FALSE, message=FALSE}
#read every file and create one dataframe per year
vb_results_muni <- map(files_muni, function(file){

  #year from filename
  year <- str_extract(basename(file), "\\d{4}")

  #read XML
  doc <- read_xml(file)

  #select municipality nodes
  municipalities <- xml_find_all(doc, ".//level[@description='Municipality']")

  #extract municipality results
  results <- map_dfr(municipalities, function(municipality){

    municipality_name <- xml_text(xml_find_first(municipality, "./labels/label[@language='nl']"))
    valid_votes <- as.numeric(xml_attr(municipality, "valid-votes"))

    #find Vlaams Belang or Vlaams Blok
    vb <- xml_find_first(municipality, "./lists/list[@party='VLAAMS BELANG' or @party='VLAAMS BLOK']")

    if (inherits(vb, "xml_missing")) {
      vb_votes <- 0
    } else {
      vb_votes <- as.numeric(xml_attr(vb, "votes"))
    }

    tibble(
      year = as.integer(year),
      municipality = municipality_name,
      vb_votes = vb_votes,
      valid_votes = valid_votes,
    )
  })

  results
})


#merge the output
vb_results_muni <- bind_rows(vb_results_muni)

#deal with the 2019 municipal mergers by merging the original municipalities
vb_results_muni <- vb_results_muni %>%
  mutate(municipality = case_when(municipality == "Kruishoutem" | municipality == "Zingem" ~ "Kruisem",
                            municipality == "Meeuwen-Gruitrode" | municipality == "Opglabbeek" ~ "Oudsbergen",
                            municipality == "Neerpelt" | municipality == "Overpelt" ~ "Pelt",
                            municipality == "Puurs" | municipality == "Sint-Amands" ~ "Puurs-Sint-Amands",
                            municipality == "Lovendegem" | municipality == "Waarschoot" | municipality == "Zomergem" ~ "Lievegem",
                            municipality == "Knesselare" ~ "Aalter",
                            municipality == "Nevele" ~ "Deinze",
                            TRUE ~ municipality)) %>%
  group_by(year, municipality) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop")

#calculate votes
vb_results_muni <- vb_results_muni %>% mutate(vb_perc = (vb_votes/valid_votes)*100) 
vb_results_muni$vb_perc <- round(vb_results_muni$vb_perc, 2)
vb_results_muni <- vb_results_muni %>% select(-valid_votes, -vb_votes)
```


# 5.2. Preparation of spatial election data

The cleaned municipal-level election results are joined to the geographical boundaries. The resulting dataset is reshaped into a wide format, with one column per election year, and exported as a GeoPackage for subsequent spatial analysis and visualisation.


```{r warning=FALSE, message=FALSE}
#join election and geographical data
muni_geom <- muni_geom %>% group_by(cd_munty_refnis, tx_munty_descr_nl,cd_rgn_refnis) %>% summarise()

munis_full <- vb_results_muni %>% left_join(muni_geom, by = c("municipality" = "tx_munty_descr_nl"))
munis_full <- munis_full %>% filter(cd_rgn_refnis == "02000") %>% st_as_sf
munis_full <- munis_full %>% select(-cd_rgn_refnis)

#pivot table
munis_plot <- munis_full %>% select(municipality, year, vb_perc, GEOMETRY)
munis_plot <- munis_plot %>% pivot_wider(names_from = year, values_from = vb_perc)

#export data
st_write(munis_plot, "output/munis_plot.gpkg")
```



# 6. Transform independent variables

The following section outlines the transforming of the original independent variables to the data structure needed for further analysis. This includes, cleaning, pre-processing and extrapolation of the data.


# 6.1. Inhabitants per municipality
```{r warning=FALSE, message=FALSE}
#read in population data and construct dataframe
population <- lapply(excel_sheets(file_inhabitants), function(sheet) {
  df <- read_excel(file_inhabitants, sheet = sheet)
  df$year <- as.numeric(sheet)
  df
}) |> bind_rows()

#deal with municipal mergers
population_long <- population %>%
  mutate(Woonplaats = case_when(Woonplaats == "Kruishoutem" | Woonplaats == "Zingem" ~ "Kruisem",
                            Woonplaats == "Meeuwen-Gruitrode" | Woonplaats == "Opglabbeek" ~ "Oudsbergen",
                            Woonplaats == "Neerpelt" | Woonplaats == "Overpelt" ~ "Pelt",
                            Woonplaats == "Puurs" | Woonplaats == "Sint-Amands" ~ "Puurs-Sint-Amands",
                            Woonplaats == "Lovendegem" | Woonplaats == "Waarschoot" | Woonplaats == "Zomergem" ~ "Lievegem",
                            Woonplaats == "Knesselare" ~ "Aalter",
                            Woonplaats == "Nevele" ~ "Deinze",
                            TRUE ~ Woonplaats)) %>%
  group_by(year, Woonplaats) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop")

#transform to long format
population_long <- population_long %>% filter(population_long$Woonplaats %in% munis_plot$municipality)
population_long <- population_long %>% rename(municipality = Woonplaats, population = Totaal)
```



# 6.2. Prosperity per municipality
```{r warning=FALSE, message=FALSE}
#pivot data to long format
prosperity_long <- prosperity %>% pivot_longer( cols = -Gemeente, names_to = "year", values_to = "prosperity" ) %>% 
                    mutate( year = as.numeric(year), prosperity = as.numeric(prosperity) ) %>% 
                    arrange(Gemeente, year)

#construct function for extrapolation of the data based on election years
estimate_municipality <- function(data) { 
  #Llinear model 
  linear_model <- lm(prosperity ~ year, data = data) 
  #years outside the observed period 
  years_to_estimate <- election_years[ !election_years %in% data$year ] 
  #predictions 
  if (length(years_to_estimate) > 0) {
    prediction_data <- data.frame(year = years_to_estimate) 
    prediction_data$linear <- predict(linear_model, newdata = prediction_data) 
    prediction_data
  } else {
    data.frame(year = numeric(0), linear = numeric(0))} 
}

#obtain estimated values for election years not included in the data
estimated_values <- prosperity_long %>% group_by(Gemeente) %>% nest() %>% 
  mutate( estimates = map( data, estimate_municipality ) ) %>% 
  select(Gemeente, estimates) %>% 
  unnest(estimates)

#merge the original data with the predictions and clean results
prosperity_long <- prosperity_long %>% filter(year %in% election_years) %>%
  select(Gemeente, year, observed = prosperity) %>% 
  full_join(estimated_values, by = c("Gemeente", "year")) %>% 
  mutate(prosperity = ifelse(!is.na(observed), observed, linear)) %>% 
  arrange(Gemeente, year)
prosperity_long <- prosperity_long %>% select(-observed, -linear)
prosperity_long <- prosperity_long %>% rename(municipality = Gemeente)
```



# 6.3. Non-EU immigrants per municipality
```{r warning=FALSE, message=FALSE}
#read in immigration data and construct data frame (for non-merged municipalities)
imm_noneu <- lapply(excel_sheets(file_imm_noneu), function(sheet) {
  df <- read_excel(file_imm_noneu, sheet = sheet)
  df$year <- as.numeric(sheet)
  df
}) |> bind_rows()

#clean data and retain only necessary information
imm_noneu <- imm_noneu %>% fill(Woonplaats)
imm_noneu <- imm_noneu %>% filter(`Nationaliteitsgroep van herkomst` == "Buiten EU27")
imm_noneu <- imm_noneu %>% mutate( Woonplaats = str_remove(Woonplaats, " \\(.*\\)"))

imm_noneu_long <- imm_noneu %>% filter(imm_noneu$Woonplaats %in% munis_plot$municipality)
imm_noneu_long <- imm_noneu_long %>% rename(municipality = Woonplaats, migrants = Totaal)


#clean data and retain on ly necessary information for merged municipalities
imi_mergers <- imi_mergers %>% filter(TX_DESCR_ORIGIN_NL != "België" & TX_DESCR_ORIGIN_NL != "Verenigd Koninkrijk" &
                      TX_DESCR_ORIGIN_NL != "Nederland" & TX_DESCR_ORIGIN_NL != "Luxemburg (Groot Hertogdom)" &
                      TX_DESCR_ORIGIN_NL != "EU-15" & TX_DESCR_ORIGIN_NL != "EU-13" &
                      TX_DESCR_ORIGIN_NL != "Duitsland" & TX_DESCR_ORIGIN_NL != "Frankrijk")
imi_mergers <- imi_mergers %>% group_by(cd_refnis_mun_25, year) %>% summarise(migrants = sum(totaal_aantal))
imi_mergers <- imi_mergers %>% filter(year < "2025")
imi_mergers$cd_refnis_mun_25 <- as.character(imi_mergers$cd_refnis_mun_25)


#select merge both data frames and retain only necessary information
imi_mergers <- imi_mergers %>% full_join(st_drop_geometry(muni_geom), by = c("cd_refnis_mun_25" = "cd_munty_refnis"))
imi_mergers <- imi_mergers %>% filter(cd_rgn_refnis == "02000")

imi_empty <- imi_mergers %>% filter(is.na(migrants))
imi_add <- imm_noneu_long %>% filter(imm_noneu_long$municipality %in% imi_empty$tx_munty_descr_nl) %>% 
  select(-`Nationaliteitsgroep van herkomst`) %>% rename(tx_munty_descr_nl = municipality)

imm_noneu_long <- bind_rows(imi_mergers, imi_add)
imm_noneu_long <- imm_noneu_long %>% select(tx_munty_descr_nl, year, migrants) %>% rename(municipality = tx_munty_descr_nl)
imm_noneu_long <- imm_noneu_long[complete.cases(imm_noneu_long$migrants),]


#extrapolate the data to account for the missing information (exponential model)
estimate_municipality <- function(data) {
  
  #first observed year for this municipality
  first_year <- min(data$year, na.rm = TRUE)
  
  #center year around first observed year
  data <- data %>%
    mutate(year_centered = year - first_year)
  
  #exponential model: migrants = a * exp(b * year_centered)
  exponential_model <- lm(log(migrants) ~ year_centered, data = data)
  
  #years outside the observed period
  years_to_estimate <- election_years[!election_years %in% data$year]
  
  #predictions
  if (length(years_to_estimate) > 0) {
    
    prediction_data <- data.frame(year = years_to_estimate) %>%
      mutate(year_centered = year - first_year)
    
    #predict on log scale and transform back
    prediction_data$exponential <- exp(predict(exponential_model, newdata = prediction_data))
    prediction_data
    
  } else {
    
    data.frame(
      year = numeric(0),
      exponential = numeric(0)
    )
  }
}

#obtain estimated values for election years not included in the data
estimated_values <- imm_noneu_long %>% group_by(municipality) %>%
  nest() %>% mutate(estimates = map(data, estimate_municipality)) %>%
  select(municipality, estimates) %>%
  unnest(estimates)

#merge the original data with the predictions and clean results
imm_noneu_long <- imm_noneu_long %>%
  filter(year %in% election_years) %>%
  select(municipality, year, observed = migrants) %>%
  full_join(estimated_values, by = c("municipality", "year")) %>%
  mutate(migrants = ifelse(!is.na(observed), observed, exponential)) %>%
  arrange(municipality, year)
```



# 6.4. Number of crimes per municipality
```{r warning=FALSE, message=FALSE}
#change data format muncipal look-up table 
muni_merged_lookup <- muni_merged_lookup %>%
  separate_rows(muni_orig, sep = ",")

#select the means of the amount of crime between election years to obtain more general trend
crime <- crime %>%
  mutate(
    `2003` = rowMeans(select(., `2000`:`2003`), na.rm = TRUE),
    `2007` = rowMeans(select(., `2004`:`2007`), na.rm = TRUE),
    `2010` = rowMeans(select(., `2008`:`2010`), na.rm = TRUE),
    `2014` = rowMeans(select(., `2011`:`2014`), na.rm = TRUE),
    `2019` = rowMeans(select(., `2015`:`2019`), na.rm = TRUE),
    `2024` = rowMeans(select(., `2020`:`2024`), na.rm = TRUE)
  ) %>%
  select(municipality, `2003`, `2007`, `2010`, `2014`, `2019`, `2024`)

crime <- crime %>% pivot_longer(cols = c(`2003`, `2007`, `2010`, `2014`, `2019`, `2024`))
crime <- crime %>% rename(year = name, crime = value)

#disaggregate data for merged municipalities based on population data and transform to long format
pop_merged <- population_long %>% filter(year == 2024 | year == 2019 | year == 2014 | year == 2010 | year == 2007 | year == 2003)
pop_merged <- pop_merged %>% left_join(muni_merged_lookup, by = c("municipality" = "muni_orig"))
pop_merged <- pop_merged[complete.cases(pop_merged$muni_merged),]
pop_merged <- pop_merged %>% group_by(muni_merged, year) %>% 
  mutate(total_population = sum(population, na.rm = TRUE),
         population_share = population / total_population) %>%
  ungroup()

crime$year <- as.numeric(crime$year)
pop_merged <- pop_merged %>% left_join(crime, by = c("muni_merged" = "municipality", "year" = "year")) %>%
  mutate(crime_disaggregated = crime * population_share)
pop_merged <- pop_merged %>% select(municipality, year, crime_disaggregated) %>% rename(crime = crime_disaggregated)

crime <- crime %>% filter(!(municipality %in% muni_merged_lookup$muni_merged))
crime_long <- crime %>% add_row(pop_merged)


#construct function for extrapolation of the data based on election years
estimate_municipality <- function(data) { 
  #linear model 
  linear_model <- lm(crime ~ year, data = data) 
  #years outside the observed period 
  years_to_estimate <- election_years[ !election_years %in% data$year ] 
  #predictions 
  if (length(years_to_estimate) > 0) {
    prediction_data <- data.frame(year = years_to_estimate) 
    prediction_data$linear <- predict(linear_model, newdata = prediction_data) 
    prediction_data
  } else {
    data.frame(year = numeric(0), linear = numeric(0))} 
}

#obtain estimated values for election years not included in the data
estimated_values <- crime_long %>% group_by(municipality) %>% nest() %>% 
  mutate(estimates = map(data, estimate_municipality ) ) %>% 
  select(municipality, estimates) %>% 
  unnest(estimates)

#merge the original data with the predictions and clean results
crime_long <- crime_long %>% filter(year %in% election_years) %>%
  select(municipality, year, observed = crime) %>% 
  full_join(estimated_values, by = c("municipality", "year")) %>% 
  mutate(crime = ifelse(!is.na(observed), observed, linear)) %>% 
  arrange(municipality, year)
crime_long <- crime_long %>% select(-observed, -linear)
crime_long <- crime_long %>% rename(municipality = municipality)
```



# 7. Prepare data for model
# 7.1. Merge independent variables

The following section includes the merging of all independent variables.

```{r warning=FALSE, message=FALSE}
canton_lookup <- canton_lookup %>%
  separate_rows(municipality, sep = ",")

municipal_data <- prosperity_long %>%
  left_join(
    population_long,
    by = c("municipality", "year")
  ) %>%
  left_join(
    imm_noneu_long,
    by = c("municipality", "year")
  ) %>%
  left_join(
    crime_long,
    by = c("municipality", "year")
  ) %>% 
  left_join(
    canton_lookup,
    by = "municipality"
  )
```



# 7.2. Aggregate independent variables at cantonal level

Aggregate all data to cantonal level and compute the final independent variables.

```{r warning=FALSE, message=FALSE}
#aggregate to canton level
canton_data <- municipal_data %>%
  group_by(canton, year) %>%
  summarise(
    prosperity = weighted.mean(
      prosperity,
      population,
      na.rm = TRUE
    ),
    migrants = weighted.mean(
      migrants,
      population,
      na.rm = TRUE
    ),
    crime = weighted.mean(
      crime,
      population,
      na.rm = TRUE
    ),
    population = sum(
      population,
      na.rm = TRUE
    ),
    .groups = "drop"
  )


#remove two unused cantons and join data for final calculations
canton_data <- canton_data %>% filter(canton != "Zaventem") %>% filter(canton != "Sint-Genesius-Rode")
canton_data <- canton_data %>% left_join(cantons_full, by = c("canton", "year"))
canton_data <- canton_data %>% st_as_sf()

#compute the final version of the independent variables
canton_data <- canton_data %>% mutate(area = as.numeric(st_area(shape)/1000000))
canton_data <- canton_data %>% mutate(migrants_perc = (migrants/population)*100)
canton_data <- canton_data %>% mutate(pop_dens = (population)/area)
canton_data <- canton_data %>% mutate(crime_rate = crime/(population/1000))
```



# 7.3. Aggregate independent variables at municipal level

Aggregate all data to municipal level and compute the final independent variables.

```{r warning=FALSE, message=FALSE}
#only select data for 2014, 2019 and 2024
municipal_data_rec <- municipal_data %>% filter(year == "2014" | year == "2019" | year == "2024")
municipal_data_rec <- municipal_data_rec %>% left_join(munis_full, by = c("municipality", "year"))
municipal_data_rec <- municipal_data_rec %>% st_as_sf()

#compute the final version of the independent variables
municipal_data_rec <- municipal_data_rec %>% mutate(area = as.numeric(st_area(GEOMETRY)/1000000))
municipal_data_rec <- municipal_data_rec %>% mutate(migrants_perc = (migrants/population)*100)
municipal_data_rec <- municipal_data_rec %>% mutate(pop_dens = (population)/area)
municipal_data_rec <- municipal_data_rec %>% mutate(crime_rate = crime/(population/1000))
```



# 7.4. Prepare each dataset for the model

calculating "between" and "within" levels of the independent variables.

```{r warning=FALSE, message=FALSE}
#canton level calculations
canton_data <- canton_data %>%
  group_by(canton) %>%
  mutate(
    prosperity_between = mean(prosperity, na.rm = TRUE),
    prosperity_within  = prosperity - prosperity_between,

    crime_between = mean(crime_rate, na.rm = TRUE),
    crime_within  = crime_rate - crime_between,

    mig_between = mean(migrants_perc, na.rm = TRUE),
    mig_within  = migrants_perc - mig_between
    
  ) %>%
  ungroup()


#split dataset according to the three predetermined phases
canton_data_1 <- canton_data %>%
  filter(year >= 1987, year <= 2007)

canton_data_2 <- canton_data %>%
  filter(year >= 2007, year <= 2014)

canton_data_3 <- canton_data %>%
  filter(year >= 2014, year <= 2024)


#municipality level calculations
municipal_data_rec <- municipal_data_rec %>%
  group_by(municipality) %>%
  mutate(
    prosperity_between = mean(prosperity, na.rm = TRUE),
    prosperity_within  = prosperity - prosperity_between,

    crime_between = mean(crime_rate, na.rm = TRUE),
    crime_within  = crime_rate - crime_between,

    mig_between = mean(migrants_perc, na.rm = TRUE),
    mig_within  = migrants_perc - mig_between
    
  ) %>%
  ungroup()


#save final data
saveRDS(municipal_data, "output/municipal_data_rec.RDS")
saveRDS(municipal_data, "output/municipal_data_.RDS")
saveRDS(canton_data, "output/canton_data.RDS")
```



# 8. Compute models and analyse output for each phase
# 8.1. Compute models

The computation of the models for each of the phases can be found in this section, including the municipal-level model.

```{r warning=FALSE, message=FALSE}
#1987-2007
model_1 <- lmer(
  vb_perc ~
    
    prosperity_between +
    prosperity_within +

    crime_between +
    crime_within +

    mig_between +
    mig_within +

    pop_dens +

    (1 | canton),

  data = canton_data_1
)

#2007-2014
model_2 <- lmer(
   vb_perc ~

    prosperity_between +
    prosperity_within +

    crime_between +
    crime_within +

    mig_between +
    mig_within +

    pop_dens +

    (1 | canton),

  data = canton_data_2
)

#2014-2024 (at canton level)
model_3_can <- lmer(
  vb_perc ~

    prosperity_between +
    prosperity_within +

    crime_between +
    crime_within +

    mig_between +
    mig_within +

    pop_dens +

    (1 | canton),

  data = canton_data_3
)

#2014-2024 (at municipal level)
model_3_muni <- lmer(
  vb_perc ~

    prosperity_between +
    prosperity_within +

    crime_between +
    crime_within +

    mig_between +
    mig_within +

    pop_dens +

    (1 | municipality),

  data = municipal_data_rec
)
```



# 8.2. Analyse model results

Check the summary and results of each model as well as potential (multi-)collinearity.

```{r warning=FALSE, message=FALSE}
#model results
summary(model_1)
summary(model_2)
summary(model_3_can)
summary(model_3_muni)
```


```{r warning=FALSE, message=FALSE}
#check for collinearity between the independent variables
check_collinearity(model_1)
check_collinearity(model_2)
check_collinearity(model_3_can)
check_collinearity(model_3_muni)
```



# 8.3. Plot model results at cantonal level

The following sections plot the model results at canton level for each of the independent variables' coefficients.

```{r warning=FALSE, message=FALSE}
#get confidence intervals of each variables coefficient for each phase
p1 <- tidy(model_1, effects = "fixed", conf.int = TRUE) %>%
  mutate(phase = "1987–2007")

p2 <- tidy(model_2, effects = "fixed", conf.int = TRUE) %>%
  mutate(phase = "2007–2014")

p3 <- tidy(model_3_can, effects = "fixed", conf.int = TRUE) %>%
  mutate(phase = "2014–2024")

p4 <- tidy(model_3_muni, effects = "fixed", conf.int = TRUE) %>%
  mutate(phase = "2014–2024")


#merge together for the cantonal models
results <- bind_rows(p1, p2, p3)

#structure results and naming in the data frame (and plot)
results_plot <- results %>%
  filter(term != "(Intercept)") %>%
  mutate(
    variable = recode(
      term,
      prosperity_between = "Prosperity (between)",
      prosperity_within  = "Prosperity (within)",
      crime_between      = "Crime rate (between)",
      crime_within       = "Crime rate (within)",
      mig_between        = "Immigration (between)",
      mig_within         = "Immigration (within)",
      pop_dens           = "Population density"
    )
  )


results_plot <- results_plot %>%
  mutate(
    variable = factor(
      variable,
      levels = c(
        "Prosperity (between)",
        "Prosperity (within)",
        "Crime rate (between)",
        "Crime rate (within)",
        "Immigration (between)",
        "Immigration (within)",
        "Population density"
      )
    )
  )


#plot results
ggplot(results_plot, aes(x = estimate, y = phase, colour = phase)) +
  geom_vline(xintercept = 0, linetype = "dashed",) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.15, show.legend = FALSE) +
  geom_point(size = 3, show.legend = FALSE) +
  
  facet_wrap(~ variable, scales = "free_x", ncol = 2) +
  labs(x = "Fixed-effect estimate (95% CI)", y = NULL, colour = "Phase") +
  theme_minimal() +
  theme(panel.spacing.x = unit(3, "lines"))
```



# 8.4. Plot model results for final phase

The following sections plot the model results for each of the independent variables' coefficients at canton level and at municipal level for the 2014-2024 phase.

```{r warning=FALSE, message=FALSE}
#merge the model results for the last phase together (both canton and municipal)
p4 <- p4 %>% mutate(type = "Municipality")
p3 <- p3 %>% mutate(type = "Canton")
  
results2 <- bind_rows(p3, p4)


#structure results and naming in the data frame (and plot)
results_plot2 <- results2 %>%
  filter(term != "(Intercept)") %>%
  mutate(
    variable = recode(
      term,
      prosperity_between = "Prosperity (between)",
      prosperity_within  = "Prosperity (within)",
      crime_between      = "Crime rate (between)",
      crime_within       = "Crime rate (within)",
      mig_between        = "Immigration (between)",
      mig_within         = "Immigration (within)",
      pop_dens           = "Population density"
    )
  )


results_plot2 <- results_plot2 %>%
  mutate(
    variable = factor(
      variable,
      levels = c(
        "Prosperity (between)",
        "Prosperity (within)",
        "Crime rate (between)",
        "Crime rate (within)",
        "Immigration (between)",
        "Immigration (within)",
        "Population density"
      )
    )
  )


#plot results
ggplot(results_plot2, aes(x = estimate, y = type, colour = type)) +
  geom_vline(xintercept = 0, linetype = "dashed",) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.15, show.legend = TRUE) +
  geom_point(size = 3, show.legend = TRUE) +
  
  facet_wrap(~ variable, scales = "free_x", ncol = 2) +
  labs(x = "Fixed-effect estimate (95% CI)", y = NULL, colour = "type") +
  theme_minimal() +
  theme(panel.spacing.x = unit(3, "lines"))
```



# 8.5. Compare cantonal and municipal model for final phase

This final sections check whether the difference in the coefficient for each of the independent variables in both models are statistically significant or not.

```{r warning=FALSE, message=FALSE}
#aggregate model data for municipal and canton level of final phase
canton_data_3$level_id <- paste0("C_", canton_data_3$canton)
municipal_data_rec$level_id <- paste0("M_", municipal_data_rec$municipality)

canton_data_3$level <- "Canton"
municipal_data_rec$level <- "Municipality"

combined_data <- bind_rows(canton_data_3, municipal_data_rec)
combined_data$level <- factor(combined_data$level, levels = c("Canton", "Municipality"))
```


```{r warning=FALSE, message=FALSE}
#run combined data model
model_level <- lmer(
  vb_perc ~
    level * (
      prosperity_between + prosperity_within +
      crime_between + crime_within +
      mig_between + mig_within +
      pop_dens
    ) +
    (1 | level_id),
  data = combined_data
)

#plot results
summary(model_level)
```

